WASP-4 is Accelerating Towards the Earth

### 2020-01-30 paper/ms.tex
fix math error in dP/dt vs baseline scaling

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Proposed explanations for the period change include tidal orbital
  decay, orbital precession, and light-travel time effects.  We
  present new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  The implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  Combining the radial velocities with new limits from
  speckle imaging, we find that the system's acceleration is likely
  caused by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis
  between 10-100$\,$AU.  The statistics from
  \citet{knutson_friends_2014} imply that about 1 in 6 hot Jupiters
  are expected to show similar period decreases as WASP-4, due to
  acceleration by outer companions.  We find that these period
  decreases will be increasingly measureable in coming years, because
  the precision of period derivative measurements scales as the fourth
  power of the observing baseline.  Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

Indirect studies of age and angular momentum indicators including the
hot Jupiter semi-major axis distribution, host star spin rates, and
galactic velocity dispersions have led to estimates for the inspiral
timescale that vary from much less to much greater than the
main-sequence evolution time  ({\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).  Direct measurements of tidal
orbital decay through transit and occultation timing could provide an
empirical resolution.  For instance, combined transit timing and
radial velocity measurements for WASP-12b have shown that its secular
period decrease of $\approx$30 milliseconds per year is almost
certainly due to tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

@@

It is therefore pertinent to ask at what point in time further
detections of the light-travel time effect will become routine for hot
Jupiters.  This question is the same as asking: at what rate does the
uncertainty in the quadratic term of Equation~\ref{eq:quadratic_model}
scale with the observing baseline?  We take a Fisher analysis approach
to the problem.  First, we rewrite Equation~\ref{eq:quadratic_model}
as
\begin{align}
  t_{\rm tra} = a_0 + a_1 E + a_2 E^2,
\end{align}
where $a_0\equiv t_0$, $a_1\equiv P$, and $a_2\equiv 0.5\cdot {\rm
d}P/{\rm d}E$.  By following \citet{gould_chi2_2003}, we derived that
if $N$ transit timing measurements are taken uniformly across a
baseline of $\Delta E$ epochs with constant precision $\sigma_{\rm
tra}$, then the uncertainty of the quadratic term is given by
\begin{align}
  \sigma_{a_2} =
  \left(\frac{25{,}920}{N}\right)^{1/2}
  \frac{\sigma_{\rm tra}}{(\Delta E)^4}
\end{align}
This result implies that a doubled observing baseline yields a
sixteen-fold improvement in precision on d$P$/d$t$.  If regular
observations continue from ground and space-based observatories,
period derivatives with precision below $1~{\rm ms}\,{\rm yr}^{-1}$
will be measureable before 2030.

@@

Most hot Jupiters have outer companions with masses larger than
Jupiter beyond 5$\,$AU
\citep{knutson_friends_2014,bryan_statistics_2016}. The accelerations
and period changes induced by these outer companions will become an
increasingly large nuisance in the hunt for orbital decay as the
observational baselines get longer.  In particular, the precision with
which the period derivative can be measured from transits scales as
the fourth power of the baseline duration (Section~\ref{sec:fisher}), and
so within a decade many more hot Jupiters should show orbital period
changes due to accelerations induced by their outer companions.  
To distinguish this effect from tidal decay, further long-term radial
velocity measurements of hot Jupiters are required.

>>>>

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Proposed explanations for the period change include tidal orbital
  decay, orbital precession, and light-travel time effects.  We
  present new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  The implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  Combining the radial velocities with new limits from
  speckle imaging, we find that the system's acceleration is likely
  caused by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis
  between 10-100$\,$AU.  The statistics from
  \citet{knutson_friends_2014} imply that about 1 in 6 hot Jupiters
  are expected to show similar period decreases as WASP-4, due to
  acceleration by outer companions.  We find that these period
  decreases will be increasingly measureable in coming years, because
  the precision of period derivative measurements scales as
  (observing baseline)$^{5/2}$.  Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

Indirect studies of age and angular momentum indicators including the
hot Jupiter semi-major axis distribution, host star spin rates, and
galactic velocity dispersions have led to estimates for the inspiral
timescale that vary from much less to much greater than the
main-sequence evolution time  ({\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).  Direct measurements of tidal
orbital decay through transit and occultation timing could provide an
empirical resolution.  For instance, combined transit timing and
radial velocity measurements for WASP-12b have shown that its secular
period decrease of $\approx$30 milliseconds per year is probably due
to tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

@@

It is therefore pertinent to ask at what point in time further
detections of the light-travel time effect will become routine for hot
Jupiters.  This question is the same as asking: at what rate does the
uncertainty in the quadratic term of Equation~\ref{eq:quadratic_model}
scale with the observing baseline?  We take a Fisher analysis approach
to the problem.  First, we rewrite Equation~\ref{eq:quadratic_model}
as
\begin{align}
  t_{\rm tra} = a_0 + a_1 E + a_2 E^2,
\end{align}
where $a_0\equiv t_0$, $a_1\equiv P$, and $a_2\equiv 0.5\cdot {\rm
d}P/{\rm d}E$.  By following \citet{gould_chi2_2003}, we derived that
if $N$ transit timing measurements are taken uniformly across a
baseline of $\Delta E$ epochs with constant precision $\sigma_{\rm
tra}$, then the uncertainty of the quadratic term is given by
\begin{align}
  \sigma_{a_2} = 6\sqrt{5}
   \frac{\sigma}{N^{1/2} (\Delta E)^2} \propto (\Delta E)^{-5/2}.
\end{align}
This result implies that a doubled observing baseline yields an
$\approx$5.7-fold improvement in precision on d$P$/d$t$.  If regular
observations continue from ground and space-based observatories,
period derivatives with precision below $1~{\rm ms}\,{\rm yr}^{-1}$
will become much more common over the coming decade.

@@

Most hot Jupiters have outer companions with masses larger than
Jupiter beyond 5$\,$AU
\citep{knutson_friends_2014,bryan_statistics_2016}. The accelerations
and period changes induced by these outer companions will become an
increasingly large nuisance in the hunt for orbital decay as the
observational baselines get longer.  In particular, the precision with
which the period derivative can be measured from transits scales with
the baseline duration to the power of 5/2 (Section~\ref{sec:fisher}), and
so within a decade many more hot Jupiters should show orbital period
changes due to accelerations from their outer companions.  
To distinguish this effect from tidal decay, further long-term radial
velocity measurements of hot Jupiters are required.
